---
title: "Script4_DE Analysis Using EdgeR"
author: "Nickie Safarian"
date: "6/14/2022"
output:
  html_document: default
  pdf_document: default
---

## R Markdown

This script contains codes for analyzing Bulk RNA-seq data using edgeR package. Both general approach and GLM modeling are presented and the results from each method are compared at the end. 

```{r, message=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install(version = "3.15")

# BiocManager::install("edgeR")
# BiocManager::install("Limma")
# BiocManager::install("Glimma")
# install.packages('Rcpp')


```

## After you have installed edgeR, you load it like normal.
```{r pressure, message=FALSE, include=FALSE}

library(BiocManager)
library(edgeR)
library(tidyverse)
library(dplyr)
library(limma)
library(Glimma)
library(Rcpp)

```

## Input data
```{r}

counts <-  readRDS("/external/rprshnas01/kcni/nsafarian/Toker_PD_Project/pd_complex-i_stratification/Data/CountMatrix.Rds")

metadataFinal <- read.csv("/external/rprshnas01/kcni/nsafarian/Toker_PD_Project/pd_complex-i_stratification/Data/Nickie/DESeq2/metadataFinal.csv")

```


## set rownames
```{r}

rownames(metadataFinal) <- metadataFinal$RNAseq_id_ParkOme2
 
```

## Prepare the count matrix
```{r}
# convert the count matrix into a dataframe and round the counts
cts <- counts %>% as.data.frame %>% round()

# only keep samples that have input in the metadataFinal
cts.PD <- cts[ , metadataFinal$RNAseq_id_ParkOme2]
 
```

## Make sure that the columns of the count data are in the same order as rows names of the metadata*
```{r}

all(colnames(cts.PD) == rownames(metadataFinal)) # same order check
all(colnames(cts.PD) %in% rownames(metadataFinal)) 
                # same samples across two data

```

## Make DGE object
### *1) Scale numerical covariates*
```{r}

metadataFinal$Age = scale(metadataFinal$Age)
metadataFinal$PMI = scale(metadataFinal$PMI)
metadataFinal$DV200 = scale(metadataFinal$DV200)
metadataFinal$PropPos = scale(metadataFinal$PropPos)

#1) Specifiy Age as a numeric covariate
Age <- scale(metadataFinal$Age)
#2) 
PMI <- scale(metadataFinal$PMI)

```

### *2)Factor categoriacal covariates*
```{r}

#1) group by Mt.types
group <- factor(metadataFinal$MT.Grouping, levels = c("NonMT_PD", "MT_PD"))

#2) Specify sex for each of the samples
Sex <- factor(metadataFinal$Sex, levels = c("M", "F"))

#3) Specify Cohort groups for each of the samples
Cohort <- factor(metadataFinal$Cohort2, levels = c("Norway", "Barcelona"))

```

### *3) Make DGE object*
```{r}

d.PD <- DGEList(counts=cts.PD, group=group)
dim(d.PD) #60237 * 79

class(d.PD) #"DGEList"/ #Print the characteristics of the DGE-list object x 

```

### *4) Add covariates to the DGE object*
```{r}

# Add grouping names columns to d.PD
d.PD$samples$group <- group 

# Add other covars to the matrix
d.PD$samples$Age <- Age
d.PD$samples$PMI <- PMI
d.PD$samples$Sex <- Sex  
d.PD$samples$Cohort <- Cohort  


#Verify new columns successFinaly added by printing the samples dataframe
d.PD$samples 

```

### *5) Annotation*
```{r, include=FALSE}

library(org.Hs.eg.db)  
library(AnnotationDbi)


gene.ids <-  AnnotationDbi::select(org.Hs.eg.db, rownames(d.PD), keytype="ENSEMBL",column="SYMBOL") #60469 genes

#'select()' returned 1:many mapping between keys and columns  

# Chck for duplicated entries  
sum(duplicated(gene.ids$ENSEMBL))
#[1] 232

#remove duplicated ones
gene.ids <- gene.ids[!duplicated(gene.ids$ENSEMBL),] 

#check rownames in DGE object matches the genes ids 
identical(gene.ids$ENSEMBL,rownames(d.PD))
#[1] TRUE

d.PD$genes <- gene.ids

```

## Transform data from the raw scale
```{r}

# keep a copy of object in case something went wrong
d.PD2= d.PD

#calculate count.per.million
CPM <- cpm(d.PD) #Calculate counts per million for x
lcpm <- cpm(d.PD, log=TRUE) #Calculate log2[cpm + 2/L]

L <- mean(d.PD$samples$lib.size) * 1e-6 #Calculate mean library size 
          #(Units= x millions of reads)
M <- median(d.PD$samples$lib.size) * 1e-6 #Calculate median library size 
          #(Units= x millions of reads)

c(L, M)

# 81.10790 , 78.37889
#note that L= 81.1 M and log2(2/81.1)= log2(0.0246)=-5.3, this means a count of zero for this data maps to a log-CPM value of −5.3 after adding the prior count or offset.

```

## Filter the data
the filtering keeps genes that have count-per-million (CPM) above k in n samples, where k is determined by #min.count and by the sample library sizes and n is determined by the design matrix.

```{r}

# Gene expression = 0. Let's check how many genes in sample 59 have 
# expression =0?

# table(rowSums(d.PD$counts==0)==59) #Print numbers of genes with
#gene expression =0 across all samples (TRUE) and 
#number of genes with gene expression >0 in at least
#one sample (FALSE)/  TRUE 11773 / FALSE 48464 
 

#Filter out lowly expressed genes

#Note: Derfault filter setting = all genes 
#with >10 counts in a mimumum number of samples
#The minimum number of samples is chosen based on 
#the minimum group sample size.
keep.exprs <- filterByExpr(d.PD, group=group) #Filter out 
          #lowly expressed genes, keeping only those above
          #defined filtering thresholds

d.PD.f <- d.PD[keep.exprs, keep.lib.sizes=FALSE] #Write the new subset of data
          #above filter thresholds, stored in keep.expres, to x

dim(d.PD.f) # 25661  *  74  


#=============Another method for filtering===================
# keep <- rowSums(cpm(d.PD)>100) >= 2
# d.PD.f <- d.PD.f[keep,]
# dim(d.PD.f)

```

## Plot log-cpm before and after filtering
```{r}

lcpm.cutoff <- log2(10/M + 2/L) #Define cutoff for log-cpm values
# here lcpm.cutoff = -2.715544

# Define samples names
samplenames <- colnames(d.PD)

library(RColorBrewer)
nsamples <- ncol(d.PD)
col <- colorRampPalette(brewer.pal(8, "Set2"))(nsamples)

par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.5), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
  den <- density(lcpm[,i])
} 
lines(den$x, den$y, col=col[i], lwd=2)

lcpm.f <- cpm(d.PD.f, log=TRUE)
plot(density(lcpm.f[,1]), col=col[1], lwd=2, ylim=c(0,0.2), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
  den.f <- density(lcpm.f[,i])
} 
lines(den.f$x, den.f$y, col=col[i], lwd=2)



```

## After filtering, it is a good idea to reset the library sizes:
```{r}

d.PD.f$samples$lib.size <- colSums(d.PD.f$counts)
d.PD.f$samples

```

## Normalizing the data
```{r}

# Normalization is required to ensure that the expression distributions of each sample are similar across the entire experiment.
d.PD.f <- calcNormFactors(d.PD.f, method = "TMM") 
          #Normalize data by Trimmed Mean M-values (TMM) 


# to access the values run: d.PD.f$samples$norm.factors

```

## Data exploration
### *Unsupervised clustering of samples*
First generate a multi-dimensional scaling (MDS) plot to conduct unsupervised clustering of samples **before filtering**.
```{r}

plotMDS(d.PD.f, method="bcv", col=as.numeric(d.PD.f$samples$group))
legend("topright" , as.character(unique(d.PD.f$samples$group)), col=1:3, pch=20)

#Note that the different types are supposed to separate out nicely.

```

Next, generate a multi-dimensional scaling (MDS) plot to conduct unsupervised clustering of **filtered samples**. 
```{r}
#Dimensions 1 and 2 are examined first using colour grouping defined by MT.types

lcpm.f <- cpm(d.PD.f, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  c("grey", "seagreen")
col.group <- as.character(col.group)
col.sex <- Sex
levels(col.sex) <-  c("pink", "blue")
col.sex <- as.character(col.sex)
plotMDS(lcpm.f, labels=group, col=col.group)
title(main="A. Sample MT.groups")
plotMDS(lcpm.f, labels=Sex, col=col.sex, dim=c(1,2))
title(main="B. Samples Sex")

#Dimensions 3 and 4 are examined next using dim()

```

## Estimating the Dispersion
```{r}
d.PD.f <- estimateCommonDisp(d.PD.f, verbose=T)
names(d.PD.f)
d.PD.f$common.dispersion # 0.1704014


#For routine differential expression analysis, we use empirical Bayes tagwise dispersions. After common dispersion was calculated run:
d.PD.f <- estimateTagwiseDisp(d.PD.f)
names(d.PD.f)

```

## Plot the tagwise biological coefficient of variation (square root of dispersions) against log2-CPM.
```{r}

plotBCV(d.PD.f)

```

## GLM estimates of dispersion 
Generalized linear models (GLMs) are an extension of classical linear models to non-normally distributed response data. It allows for more complex designs.
### *Step 1)Design a model matrix*

```{r}
# Method 1:
# We need to adjust for baseline differences between the patients, so the 
# first step is to initialize the design matrix with sample effects:
# design <- model.matrix(~ 0+group)
# 
# #Then we define disease-specific treatment effects and append them to the design matrix:
# design <- cbind(design, Age, PMI, Cohort, Sex)
# colnames(design) <- gsub("group", "", colnames(design))
# head(design)

#===============================================================
## Method 2:
# We need to adjust for baseline differences between the patients, so the first step is to initialize the design matrix with patient effects:
design <- model.matrix(~ 0+ group + Sex + group*Sex + Age+ PMI+ Cohort)


rownames(design) <- colnames(d.PD)

#Then we define disease-specific treatment effects and append them to the design matrix:
colnames(design) <- gsub("group", "", colnames(design))
colnames(design)

```

### *Step2) Define contrasts for pairwise comparisons*
```{r}

# Note : You can define contrast first then fit a model or you can skip designing the contarst matrix at this step.

# contr.matrix <- makeContrasts(
#   NonMT.vs.MT = NonMT_PD - MT_PD,  #????? Doesn't work, DONT KNOW WHY?
#   levels = colnames(design))
# 
# 
# contr.matrix 

```

### *Step3) Estimate GLM Dispersion*
```{r}

d.PD.f <- estimateGLMCommonDisp(d.PD.f,design)
d.PD.f <- estimateGLMTrendedDisp(d.PD.f,design, method="power")
# You can change method to "auto", "bin.spline", "power", "spline", 
# "bin.loess". The default is "auto" which chooses "bin.spline" when > 200 tags # and "power" otherwise.


d.PD.f <- estimateGLMTagwiseDisp(d.PD.f,design)
plotBCV(d.PD.f)

```

## Testing for differential expression (DE)
Once the dispersion is estimated, we can proceed with detecting differential expression patterns.

## *Method 1:General Approach*
```{r}
et.12 <- exactTest(d.PD.f, pair=c(1,2)) # compare groups NMT_PD and MT_PD

topTags(et.12, n=10)


```

The total number of differentially expressed genes at FDR< 0:05 is:
```{r}

de.1 <- decideTestsDGE(et.12, adjust.method="BH", p.value=0.05)

summary(de.1) # Down 6317 /  Up 7439

```

Here the entries for -1, 0 and 1 are for down-regulated, non-differentially expressed and up-regulated tags respectively.

```{r}

de1tags12.PD <- rownames(d.PD.f)[as.logical(de.1)] 

plotSmear(et.12, de.tags=de1tags12.PD)  # generates a plot
abline(h = c(-1, 1), col = "blue")      # of the tagwise log-fold-changes                                         # against log-cpm                     
```


### save the sigs.results
```{r}
et.12.df <- as.data.frame(et.12)
rownames(et.12.df) <- et.12.df$ENSEMBL
et.12.df.sigs <- et.12.df[rownames(d.PD.f)[as.logical(de.1)] , ]

write.csv(et.12.df, "DEG.all.Results.EdgeR.general.method.csv")
write.csv(et.12.df, "DEG.sigs.Results-padj0.05-EdgeR.general.method.csv")
```

## *Method 2: GLM testing for differential expression*
Just as we used a GLM to fit the trend line above, we can also use this in finding the tags that are interesting by using a likelihood ratio test.
```{r}

# generate a second copy of DGE object
d.PD.f2 =d.PD.f

# to fit a model, GLM uses the defined design  
fit <- glmFit(d.PD.f2, design)

# compare (group 1 - group 2) to 0:
# this is equivalent to comparing group 1 to group 2
lrt1.2 <- glmLRT(fit, contrast=c(1,-1, 0, 0, 0, 0, 0))

topTags(lrt1.2, n=10)

#Note that glmLRT has conducted a test for the last coefficient in the linear model, which we can see is the Sex effect:
#  colnames(design)
# "NonMT_PD"  "MT_PD"   "SexF"   "Age"   "DV200"          
# "CohortBarcelona"     "MT_PD:SexF" 

```

```{r}

de.2 <- decideTestsDGE(lrt1.2, adjust.method="BH", p.value = 0.05)
de2tags12.PD <- rownames(d.PD.f2)[as.logical(de.2)]
plotSmear(lrt1.2, de.tags=de2tags12.PD)
abline(h = c(-2, 2), col = "blue")

```

### save the sigs.results
```{r}
Irt1.2.df <- as.data.frame(lrt1.2)
rownames(Irt1.2.df) <- Irt1.2.df$ENSEMBL
Irt1.2.df.sigs <- Irt1.2.df[rownames(d.PD.f2)[as.logical(de.2)] , ]

write.csv(Irt1.2.df, "DEG.all.Results.GlmLRT.method.csv")
write.csv(Irt1.2.df.sigs, "DEG.sigs.Results-padj0.05-GlmLRT.method.csv")
```

## *Method 3: GLMQL testing for differential expression*
```{r}

fit.2 <- glmQLFit(d.PD.f2, design)

plotQLDisp(fit.2)

```

```{r}

qlf.Mito <- glmQLFTest(fit.2, contrast=c(-1,1,0,0,0,0,0)) 
                       #Coefficient:  -1*NonMT_PD 1*MT_PD

# another way to set this contrast is to use coef number:
# qlf <- glmQLFTest(fit, coef=1: 2) #1 is NonMT_PD and 2 is MT_PD

```


```{r}

# To find genes that express differentially based on interaction between sex and Mito-type covariates*
qlf.Sex.intr.Mito <- glmQLFTest(fit.2, coef=7)# Coefficient:  "MT_PD:SexF"  

```


```{r}

# Summary of sig. results for mito-types
topTags(qlf.Mito)
FDR <- p.adjust(qlf.Mito$table$PValue, method="BH")
sum(FDR < 0.05)  #10258
summary(decideTests(qlf.Mito))

```

```{r}

# Summary of sig. results for the interaction term
topTags(qlf.Sex.intr.Mito)
FDR <- p.adjust(qlf.Sex.intr.Mito$table$PValue, method="BH")
sum(FDR < 0.05)  
summary(decideTests(qlf.Sex.intr.Mito))

```

### save the sigs.results
```{r}
qlf.Mito.df <- as.data.frame(qlf.Mito)
write.csv(qlf.Mito.df, "DEG.Results.glmQLFit.method.csv")

```


## *Method 4: use Limma to generate linear model and voom plot without voom precision weights*
```{r}

library(limma) #Load limma 
library(edgeR) #Load edgeR

# x is the DGE object , here we called it d.PD
#function
v <- voom(d.PD.f, design, plot=TRUE)

```

### Plots of variance vs. mean here will now
```{r}

#have voom-generated precision weights incorporated into the generation of the plot

vfit <- lmFit(v, design=design)
vfit <- contrasts.fit(vfit, contrast=c(-1,1,0,0,0,0,0)) 
efit <- eBayes(vfit)
plotSA(efit)

```

### Examine the number of differentially expressed genes 
```{r}

#Print out a summary of the numbers of upregulated
          #and downregrulated genes using:
          #a) P-value cutoff (default P<0.05)

summary(decideTests(efit))  # shows 0 genes

```

### Print out a summary of the numbers of upregulated and downregrulated genes using:
```{r}
         
#a) P-value cutoff (default P<0.05)
#b) minimum log-FC requirement --> lfc=0.25

tfit <- treat(vfit, lfc=0.25) 
dt <- decideTests(tfit) 
summary(dt)

```

### List the most significantly differentially expressed genes (smallest to largest adjusted P) 
```{r}
          #from results of differential 
          #expression analysis using the "eBayes" function
          

NonMT.vs.MT <- topTable(efit, coef=1, n=Inf) 
 
head(NonMT.vs.MT)


```

### save the results
```{r}

NonMT.vs.MT.sigs <- NonMT.vs.MT %>% subset(NonMT.vs.MT$adj.P.Val <0.05)

write.csv(NonMT.vs.MT.sigs, "DEG.sigs.Results.Limma.Voom.method.csv")

```


### GRAPHICALLY REPRESENT DIFFERENTIAL EXPRESSION ANALYSIS DATA             
```{r}
#Mean Difference (MD) Plot 
#(logFC vs. log-cpm)

#Note: Differentially expressed genes are highlighted
#in the plot.

plotMD(tfit, column=1, status=dt[,1], main=colnames(tfit)[1],
       xlim=c(-8,13))

```

## Heatmap
```{r,message=FALSE, error=FALSE}

# library(gplots)
# 
# NonMT.vs.MT.topgenes <- NonMT.vs.MT$ENSEMBL[1:100]
# i <- which(v$genes$ENSEMBL %in% NonMT.vs.MT.topgenes)
# mycol <- colorpanel(1000,"blue","white","red")
# heatmap.2(lcpm.f[i,], scale="row",
#           labRow=v$genes$SYMBOL[i], labCol=group,
#           col=mycol, trace="none", density.info="none",
#           margin=c(0.5,0.5),
#           dendrogram="column")


# If an error poped for the margins being too large , do this:
# in console type par("mar") --> it'll show 5.1 4.1 4.1 2.1, 
# now, set the image margin to this par(mar=c(0.1,0.1,0.1,0.1))

# If that didn't worked use dev.off()

# If dev.off returned errors use, dev.set(dev.next())

```



